Discrete Ordinates

### 2017-07-05 docprep/chapimpl.tex
Fixed typos and updated LaTeX template

This chapter covers the details of the implementation of the different components of the DOCTORS code base. The code is available from a GitHub Git repository. At the time of this writing, the repository is located at \texttt{https://github.com/eNorris/thesis} and is only available subject to special request and license agreement. An executable version is available.

@@

INSERT NOTE ABOUT CODE LISTING FORMATTING

@@

On the CPU, data is stored as C++ \texttt{std::vector<T>} objects where the \texttt{T} template parameter can be either \texttt{float} or \texttt{double} depending on the precision (32 bits or 64 bits respectively) needed. The storage space available to a \texttt{std::vector<T>} object is effectively unlimited, bounded only by the host CPU's physical RAM. However, other effects such as memory fragmentation can limit the practical size of a single, continuous \texttt{std::vector<T>} object significantly. To emulate indexing of a $N_x \times N_y$ matrix, a single \texttt{std::vector<T>} is created and resized to $N_x \times N_y$ elements. The global index, $i$, in the flattened array is then
\begin{equation}
i = i_x N_y + i_y.
\end{equation}
This pattern extends to multi-dimensional matrices. For example, a $N_x \times N_y$ matrix would be globally indexed as
\begin{equation}
i = i_x N_y N_z + i_y N_z + i_z.
\end{equation}

@@

\section{Cross Section Parsing}\label{sec:xsparse}
AMPX cross sections are stored in a binary format. Figure~\ref{fig:ampx} shows the data format of the overall file. The binary file is split into a sequence of records. The header section of the file has overall information about the file including the number of nuclides stored, their energy group structure, and a comment string describing file. Following the header section is a list of directory records. Each directory has information about a specific nuclide including the cross section reaction types available, temperatures for which the cross sections were evaluated, the number of Legendre expansion coefficients stored in the scatter matrix, and other data. Following the directories, more records follow that describe the energy structure used in the file. There is one such record for each particle type. The libraries used in the current version of DOCTORS have both neutron and photon cross section data stored.

@@

The header always reports the number of 8-bit bytes required to store the data. However, some data types, such as texttt{char}, are only one byte so each word represents multiple (4 in the case of \texttt{char}) distinct characters.

@@

Qt5 prvides a user interface for building user interfaces. Components such a sbuttons, drop boxes, radio buttons, etc. are made available to the user. Through extensive use of polymorphism, Qt simplifies the addition of graphical interactive elements called widgets. All widgets inherit from the base \texttt{QWidget} class which inherits from \texttt{QObject}. Any class that utilizes the signal/slot mechanism must extend the \texttt{QOjbect} class. As an example, a button can be created in the Qt5 user interface and named \texttt{button1}. This object will be accessible as \texttt{ui->button1}. When the user clicks on the button, the signal \texttt{button1.clicked()} will automatically be emitted. The user can connect this signal to any slot with an identical number arguments of the same type. The \texttt{button1.clicked()} signal can be connected to the slot \texttt{doStuff1()} but not \texttt{doStuff2(int)} since the arguments are not compatible.

Listing~\ref{lst:sync1} gives a pseudocode snippet that has a long function that will block the user interface. A corresponding dequence diagram is given in Fig.~\ref{fig:sync1}. When the user interacts with the GUI, the writer object begins executing the \texttt{doWrite()} function. Until this function completes, the UI thread will be busy an dunable to handle additionaly user interaction or updates. This results in the GUI becoming unresponsive and potentially issuing a warning to the user from the operating system.

@@

	// When a button named button1 which was created in the Qt5
	creator interface is clicked, the clicked() signal is 
	automatically emitted which executes the doSomethingLong()
	function
	connect(ui->button1, clicked(), this, 
		doSomethingLong());
}
\end{minted}
\caption{A long function causes the user interface to block.}\label{lst:sync1}
\end{listing}

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/writer_sync}
  \end{center}
  \caption{Sequence diagram for the synchronous call.}
\label{fig:sync1}
\end{figure}%

The code in Listing~\ref{lst:sync1} is updated to use the Qt signal/slot mechanism whose code is given in Listing~\ref{lst:async1}.

@@

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/writer_async}
  \end{center}
  \caption{Sequence diagram for the synchronous call.}
\label{fig:async1}
\end{figure}%

@@

INSERT PICTURES OF THE GUI

\section{CUDA}\label{sec:cuda}
CUDA code is compoiled with the Nvidia complier nvcc. Qt5 uses the gcc compiler and its MOC generator for meta code. In order to connect CUDA code to the Qt MOC, the CUDA code is compiled by nvcc to produce a .o file. Qt5 then compiles all other files into corresponding.o files. The linker than automactically picks up all of the .o files gneerated by nvcc. The final result is an executable that has a Qt5 generarted user interface that can communicat with an Nvidia graphics card through the CUDA language.

The sweep through the mesh is different.

@@

We first split the sweep process into $N_L$ subsweeps. For a $N_x \times N_x \times N_x$ mesh, the subsweep process is straightforward. Some of the initial subsweeps are shown for a $8 \times 8 \times 8$ mesh in Fig.~\ref{fig:subsweep_cube}. This is valid for any direction, $\hat{\Omega}$, for which all three of its direction cosines ($\mu$, $\xi$, and $\eta$) are positive.

In general, blah blah.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general1}
        \caption{$S=0$}
        \label{fig:subsweep_general1}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general2}
        \caption{$S=1$}
        \label{fig:subsweep_general2}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
    %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general3}
        \caption{$S=2$}
        \label{fig:subsweep_general3}
    \end{subfigure}
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general4}
        \caption{$S=3$}
        \label{fig:subsweep_general4}
    \end{subfigure}

    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general5}
        \caption{$S=4$}
        \label{fig:subsweep_general5}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general6}
        \caption{$S=5$}
        \label{fig:subsweep_general6}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
    %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general7}
        \caption{$S=6$}
        \label{fig:subsweep_general7}
    \end{subfigure}
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general8}
        \caption{$S=7$}
        \label{fig:subsweep_general8}
    \end{subfigure}

@@

The number of parallel tasks, $P$, that can be done on subsweep $S$ of an $N_x \times N_y \times N_z$ mesh is given by Eq.~\ref{eq:taskspersub}.

\begin{equation}\label{eq:taskspersub}
P = C_S - L_x - L_y - L_z + G_{xy} + G_{yz} + G_{xz}
\end{equation}

\begin{equation}
C_S = \frac{(S+1)(S+2)}{2}
\end{equation}

\begin{equation}
L_x = \frac{d_x(d_x+1)}{2}
\end{equation}

\begin{equation}
L_y = \frac{d_y(d_y+1)}{2}
\end{equation}

\begin{equation}
L_z = \frac{d_z(d_z+1)}{2}
\end{equation}

\begin{equation}
G_{xy} = \frac{d_{xy}(d_{xy}+1)}{2}
\end{equation}

\begin{equation}
G_{yz} = \frac{d_{yz}(d_{yz}+1)}{2}
\end{equation}

\begin{equation}
G_{xz} = \frac{d_{xz}(d_{xz}+1)}{2}
\end{equation}

\begin{equation}
d_x = \max(S+1-N_x, 0)
\end{equation}

\begin{equation}
d_y = \max(S+1-N_y, 0)
\end{equation}

\begin{equation}
d_z = \max(S+1-N_z, 0)
\end{equation}

\begin{equation}
d_{xy} = \max(S+1-N_x - Ny, 0)
\end{equation}

\begin{equation}
d_{yz} = \max(S+1-N_y - Nz, 0)
\end{equation}

\begin{equation}
d_{xz} = \max(S+1-N_x - Nz, 0)
\end{equation}

Each voxel in a subsweep can be computed in parallel. Mathematically, the $i^{th}$ subsweep from all directions can be computed in parallel. However, in practice, this results in a race condition on the GPU hardware.

DELETE BELOW HERE.

The number of elements before the 4$^{th}$ level is 1 + 2 + 3 = 6. This is simply the sum of all integers less than the level number given by Eq.~\ref{eq:levelsum}. This is conincidentally the index of the first element of that level. However, there is a special case of $N_0 = 1$.

\begin{equation} \label{eq:levelsum}
N_L = \sum_{i=1}^{L} = \frac{L(L+1)}{2}
\end{equation}

Given an index, $i$, the level to which it belongs can be computed by substituting $i$ for $N_L$ in Eq.~\ref{eq:levelsum} and solving for $L$. The index value is then floored.

\begin{equation} \label{eq:levelquadratic}
i = \frac{L(L+1)}{2} \rightarrow L^2 + L - 2i = 0
\end{equation}

\begin{equation} \label{eq:levelquadsol1}
L = \left \lfloor{\frac{-1 \pm \sqrt{1+8i}}{2}} \right \rfloor
\end{equation}

Since the index must be a positive value, the $\pm$ sign can be removed from Eq.~\ref{eq:levelquadsol1} yielding Eq.~\ref{eq:levelquadsol2}

\begin{equation} \label{eq:levelquadsol2}
L = \left \lfloor{\frac{-1 + \sqrt{1+8i}}{2}} \right \rfloor
\end{equation}

The maximum index of any level in a the $N$ subsweep is $N$. Each subsweep adds an additional level. Therefore, the ix, iy, and iz indices can be computed using Eq.~\ref{eq:levelix},~\ref{eq:leveliy}, and~\ref{eq:leveliz} respectively.

\begin{equation} \label{eq:levelix}
ix = N - N_L
\end{equation}

\begin{equation} \label{eq:leveliy}
iy = L + N_L - i
\end{equation}

\begin{equation} \label{eq:leveliz}
iz = L + i - N_L
\end{equation}

\section{MCNP Generation}\label{sec:mcnpgen}
DOCTORS has the capability to generate MCNP6 input files from the CT mesh data and source specification provided by the user.

\section{Hardware}\label{sec:hardware}
For this work, a computer with an Intel i7-5960X 8 core (16 hyperthreads) processor with a base clock speed of 3.X GHz and an Nvidia Titan Z graphics card was used. Currently, if the problem requires more memory than is available on the GPU, the problem can still be solved, but much more time will be required to to copy overhead between the CPU and GPU. If the problem requires more memory than either the GPU or CPU can provide, an error is thrown and the simulation is not run.

CUDA is Nvidia only so AMD is not considered.

>>>>

This chapter covers the details of the implementation of the different components of the DOCTORS code base. The code is available from a GitHub Git repository. At the time of this writing, the repository is located at \texttt{https://github.com/eNorris/thesis} and is only available subject to special request and license agreement. An executable version can also be made available upon request. An executable-only public release is planned for the future.

@@

Unless noted otherwise, all code listings are formatted as C++ code though some parts of the code may be omitted. Omissions include typical "boiler plate" code that accompanies all code such as the \texttt{\#include} statements and any \texttt{using namespace} commands. Additionally, the \texttt{main()} function and others are omitted, showing only the code relevant to the discussion.

@@

On the CPU, data is stored as C++ \texttt{std::vector<T>} objects where the \texttt{T} template parameter can be either \texttt{float} or \texttt{double} depending on the precision (32 bits or 64 bits respectively) needed. The storage space available to a \texttt{std::vector<T>} object is effectively unlimited, bounded only by the host CPU's physical RAM. However, other effects such as memory fragmentation can limit the practical size of a single, continuous \texttt{std::vector<T>} object significantly. To emulate indexing of a $N_x \times N_y$ matrix, a single \texttt{std::vector<T>} is created and resized to $N_x \times N_y$ elements. The global index, $i$, in the flattened array is then
\begin{equation}
i = i_x N_y + i_y.
\end{equation}
This pattern extends to multi-dimensional matrices. For example, a $N_x \times N_y \times N_z$ matrix would be globally indexed as
\begin{equation}
i = i_x N_y N_z + i_y N_z + i_z.
\end{equation}

@@

\section{Cross Section Parsing}\label{sec:xsparse}
AMPX cross sections are stored in a binary format. These files are \textit{not} distributed with DOCTORS but must be obtained from SCALE6.2 or another code. Figure~\ref{fig:ampx} shows the data format of the overall file. The binary file is split into a sequence of records. The header section of the file has overall information about the file including the number of nuclides stored, their energy group structure, and a comment string describing the file. Following the header section is a list of directory records. Each directory has information about a specific nuclide including the cross section reaction types available, temperatures for which the cross sections were evaluated, the number of Legendre expansion coefficients stored in the scatter matrix, and other data. Following the directories, more records follow that describe the energy structure used in the file. There is one such record for each particle type. The libraries used in the current version of DOCTORS have both neutron and photon cross section data stored.

@@

The header always reports the number of 8-bit bytes required to store the data. However, some data types, such as \texttt{char}, are only one byte so each word represents multiple (4 in the case of \texttt{char}) distinct characters.

@@

Qt5 prvides a user interface for building user interfaces. Components such as buttons, drop boxes, radio buttons, etc. are made available to the user. Through extensive use of polymorphism, Qt5 simplifies the addition of graphical interactive elements called widgets. All widgets inherit from the base \texttt{QWidget} class which inherits from \texttt{QObject}. Any class that utilizes the signal/slot mechanism must extend the \texttt{QOjbect} class. As an example, a button can be created in the Qt5 user interface and named \texttt{button1}. This object will be accessible as \texttt{ui->button1}. When the user clicks on the button, the signal \texttt{button1.clicked()} will automatically be emitted. The user can connect this signal to any slot with an identical number arguments of the same type. The \texttt{button1.clicked()} signal can be connected to the slot \texttt{doStuff1()} but not \texttt{doStuff2(int)} since the arguments are not compatible.

Listing~\ref{lst:sync1} gives a C++ snippet that has a long function that will block the user interface. A corresponding sequence diagram is given in Fig.~\ref{fig:sync1}. When the user interacts with the GUI, the writer object begins executing the \texttt{doWrite()} function. Until this function completes, the UI thread will be busy and unable to handle additional user interaction or updates. This results in the GUI becoming unresponsive and potentially issuing a warning to the user from the operating system. The code in Listing~\ref{lst:sync1} is updated to use the Qt signal/slot mechanism whose code is given in Listing~\ref{lst:async1}.

@@

	// When a button named button1 which was created in the Qt5
	//   creator interface is clicked, the clicked() signal is 
	//   automatically emitted which executes the doSomethingLong()
	//   function
	connect(ui->button1, clicked(), this, 
		doSomethingLong());
}
\end{minted}
\caption{A long function causes the user interface to block.}\label{lst:sync1}
\end{listing}

@@

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/writer_sync}
  \end{center}
  \caption{Sequence diagram for the synchronous call.}
\label{fig:sync1}
\end{figure}

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/writer_async}
  \end{center}
  \caption{Sequence diagram for the synchronous call.}
\label{fig:async1}
\end{figure}

@@

The data display plots that show materials and other geometry as well as the flux output use an implementation of the built-in \texttt{QGraphicsView} object to render a matrix of \texttt{QGraphicsRectItem}s on the screen. The data to be drawn on the screen is populated and some metrics such as the dataset minimum and maximum are computed. The data value at each point is evaluated to determine which bin it belongs to.

The first tab is the geometry input, shown in Fig.~\ref{fig:gui_tab1}. In the current version of DOCTORS, the only supported format is unsigned 16-bit binary files whose size is known \textit{a priori}. When the user clicks on the "Open" button, the user is prompted with a dialog for selection of a binary file. The file is read in as a list of $N_x \times N_y \times N_z$ CT numbers. As soon as the data is read in and some checks are performed, the user is prompted for selection of a CT-number-to-material conversion. Currently, two conversions are included in DOCTORS. The first is a water/air only conversion and the second is a more complex conversion to a series of dosimetrically equivalent materials representative of a human patient. Once the user specifies a conversion, the geometry explorer becomes available which allows the user to visualize the geometry file to verify that it was read and interpreted correctly.

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/gui3}
  \end{center}
  \caption{The geometry selection tab. The user must first specify the dimensionality of the voxel phantom and then load it from a binary file.}
\label{fig:gui_tab1}
\end{figure}

Some screenshots of the geometry explorer are shown in Fig.~\ref{fig:gui_geom}. The geometry viewer can show slices through the voxel phantom along all three major planes. The depth of the slice viewed is changed by moving the scroll bar in the center of the viewer up or down. As the scroll bar moves, the number at its bottom is updated accordingly to indicate the depth (in voxels) of the slice. In addition to the material, the geometry viewer can render the physical density (in g/cm$^3$) and the atom density (in atom/b-cm) as well as the raw CT number used to generate the material and density.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/gui1}
        \caption{}
        \label{fig:gui1}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs/gui4}
        \caption{}
        \label{fig:gui4}
    \end{subfigure}
    \caption{The geometry explorer. (a) The material viewer which shows which material each voxel was converted to. (b) The atom density plot.}\label{fig:gui_geom}
\end{figure}

The next three tabs identify the cross section dataset to use for the material generation, the quadrature, and the solution anisotropy treatment used. All three tabs are relatively strightforward and contain few widgets. The final tab, shown in Fig.~\ref{fig:gui_src} defines the source. The user is presented with a number of different source geometries to choose from. Each type is described more fully in Sec.~\ref{sec:source_gen}. Clicking on the "Energy Distribution" button will reveal the popup shown in Fig.~\ref{fig:gui_spect}. This popup cannot be fully populated until \textit{after} the cross section is loaded in the "Cross Section" tab since the cross section data file defines the energy group structure.

The energy distribution dialog, shown in Fig.~\ref{fig:gui_spect}, can plot the $x$ axis on either a linear or logarithmic scale. The user can click and drag to "paint" a spectrum to generate quick, qualitative spectra suiting the user's needs. Alternatively, the user can select from any of the preset spectra in the dropdown menu. Currently, only the uniform beam and diagnostic pulse distributions described later are present.

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/gui2}
  \end{center}
  \caption{The source tab. The user must select a geometry type and enter all applicable values as well as define the source energy spectrum.}
\label{fig:gui_src}
\end{figure}

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/gui6}
  \end{center}
  \caption{The spectrum energy distribution. The spectrum can be displayed on either a linear or logarithmic scale on the $x$ axis. The user can either click and drag on the spectrum to "paint" or select a preset.}
\label{fig:gui_spect}
\end{figure}

Once all necessary data is loaded, the "Launch" button becomes active and will remain so until the user creates a conflicting set of input that would prevent the solver from being able to run. When the user clicks on the launch button, the raytracer begins running asynchronously and the output display shown in Fig.~\ref{fig:gui_out} opens automatically. When the raytracer completes, the output dialog is updated with the flux information and the solver begins executing asynchronously. As each iteration completes, the output dialog is progressively updated so that the user can monitor the evolution of the solution.

The output dialog has controls very similar to the geometry explorer. The user can freely choose the plane to slice through the solution and can plot the solution on either a linear or logarithmic scale. The "Distribution" radio buttons allows the user to choose whether the uncollided, collided, or total flux is shown. The "Level Local Scale" checkbox allows the user to control whether the current slice colormap is scaled to the contents of the viewed slice or the contents of the entire voxel phantom. Checking the "Level Local Scale" box will ensure that the viewed data is normalized to include only the viewed data. Unchecking the box will scale the viewed data based on the range of the entire voxel phantom.

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/gui5}
  \end{center}
  \caption{The output dialog. The controls are similar to those found in the geometry viewer.}
\label{fig:gui_out}
\end{figure}

In the current version of DOCTORS, dosimetric output is only written to an output file and cannot be visualized graphically. The output format is designed to be human readable and easily read by any major programming language. Listing~\ref{lst:output_fmt} gives the code to read the file format where the \texttt{prod()} algorithm is given in Listing~\ref{lst:prod}. Note that \texttt{LIST} may be any appropriately typed list object such as a \texttt{std::vector} and a \texttt{LIST2D} is a list of lists or vector of vectors.

\begin{listing}
\begin{minted}[frame=lines,linenos]{c++}
// Declare variables
int N;
LIST S;
LIST2D dims;
LIST data;

// Read the number of dimensions
READ_INT(N);

// Read the size of each dimension
for(int i = 0; i < N; i++)
	READ_INT(S[i]);

// Read the axis labels
for(int i = 0; i < N; i++)
	for(int j = 0; j < S[i]; j++)
		READ_FLOAT(dims[i][j]);

// Read the data
int dataElem = prod(S);
for(i = 0; i < dataElem; i++)
	READ_FLOAT(data[i]);
\end{minted}
\caption{Algorithm to read the output format used by DOCTORS.}\label{lst:output_fmt}
\end{listing}

\begin{listing}
\begin{minted}[frame=lines,linenos]{c++}
int prod(LIST p)
{
	int product = 1;
	for(int i = 0; i < p.SIZE(); i++)
		product = product * p[i];
	return product;
}
\end{minted}
\caption{The \texttt{prod()} algorithm}\label{lst:prod}
\end{listing}

\section{CUDA}\label{sec:cuda}
CUDA code is compiled with the Nvidia complier \texttt{nvcc}. Qt5 uses the \texttt{gcc} compiler and its MOC generator for meta code. In order to connect CUDA code to the Qt MOC, the CUDA code is compiled by \texttt{nvcc} to produce an object (\texttt{.o}) file. Qt5 then compiles all other files into corresponding object files. The linker then automactically picks up all object files generated by \texttt{nvcc}. The final result is an executable that has a Qt5 generated user interface that can communicate with an Nvidia graphics card through the CUDA language.

In order to accelerate the discrete ordinate solution on a GPU, the concurrent tasks at any given moment must be known. In the single-threaded version of the code, the solver sweeps through the voxels one by one in a pre-determined fashion based on the direction. At the very beginning, the input flux to a single voxel is known from its boundary conditions. However, once that voxel's flux is computed, all three of its outgoing flux values enable three voxels to be computed independently of each other. After those three, six can be computed. Each layer of voxels whose flux can be computed independently is called a "subsweep." Figure~\ref{fig:subsweep_cube} illustrates some subsweeps through a cubic mesh.

In order to parallelize the sweep through the mesh, the global index of each voxel that can be computed in the $S$ subsweep must be known. When those indices are known, each voxel in subsweep $S$ can be solved independently by a CUDA kernel. This task is greatly simplified by noticing that the $x$, $y$, and $z$ indices of all voxels in the fourth subsweep, shown in Fig.~\ref{fig:subsweep_cube} all sum to 4 as shown in Table~\ref{table:subsweep}.

@@

The special case of the cube shown in Fig.~\ref{fig:subsweep_cube} can be extended to a more general case illustrated in Fig.~\ref{fig:subsweep_general}. Notice that the aforementioned intuition that $i_x + i_y + i_z = S$ is not necessarily true. The final subsweep ($S = 14$) contains a single voxel even though many combinations of three integers will sum to 14. The additional constraint is
\begin{equation}\label{eq:constraints}
\begin{split}
i_x &< N_x \\
i_y &< N_y \\
i_z &< N_z.
\end{split}
\end{equation}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general1}
        \caption{$S=0$}
        \label{fig:subsweep_general1}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general2}
        \caption{$S=1$}
        \label{fig:subsweep_general2}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general3}
        \caption{$S=2$}
        \label{fig:subsweep_general3}
    \end{subfigure}
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general4}
        \caption{$S=3$}
        \label{fig:subsweep_general4}
    \end{subfigure}

    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general5}
        \caption{$S=4$}
        \label{fig:subsweep_general5}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general6}
        \caption{$S=5$}
        \label{fig:subsweep_general6}
    \end{subfigure}
    ~ 
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general7}
        \caption{$S=6$}
        \label{fig:subsweep_general7}
    \end{subfigure}
    \begin{subfigure}[b]{0.2\textwidth}
        \includegraphics[width=\textwidth]{figs/subsweep_general8}
        \caption{$S=7$}
        \label{fig:subsweep_general8}
    \end{subfigure}

@@

The constraints listed in Eq.~\ref{eq:constraints} can be used to compute the total number of parallel tasks in any subsweep. The number of parallel tasks, $P$, that can be done on subsweep $S$ of an $N_x \times N_y \times N_z$ mesh is given by
\begin{equation}\label{eq:taskspersub}
P = C_S - L_x - L_y - L_z + G_{xy} + G_{yz} + G_{xz}
\end{equation}
where $C_S$, $L_x$, $L_y$, $L_z$, $G_{xy}$, $G_{yz}$, and $G_{xz}$ are defined by Equations \ref{eq:cs}-\ref{eq:dxz}.
\begin{equation}\label{eq:cs}
C_S = \frac{(S+1)(S+2)}{2}
\end{equation}
\begin{equation}
L_x = \frac{d_x(d_x+1)}{2}
\end{equation}
\begin{equation}
L_y = \frac{d_y(d_y+1)}{2}
\end{equation}
\begin{equation}
L_z = \frac{d_z(d_z+1)}{2}
\end{equation}
\begin{equation}
G_{xy} = \frac{d_{xy}(d_{xy}+1)}{2}
\end{equation}
\begin{equation}
G_{yz} = \frac{d_{yz}(d_{yz}+1)}{2}
\end{equation}
\begin{equation}
G_{xz} = \frac{d_{xz}(d_{xz}+1)}{2}
\end{equation}
\begin{equation}
d_x = \max(S+1-N_x, 0)
\end{equation}
\begin{equation}
d_y = \max(S+1-N_y, 0)
\end{equation}
\begin{equation}
d_z = \max(S+1-N_z, 0)
\end{equation}
\begin{equation}
d_{xy} = \max(S+1-N_x - Ny, 0)
\end{equation}
\begin{equation}
d_{yz} = \max(S+1-N_y - Nz, 0)
\end{equation}
\begin{equation}\label{eq:dxz}
d_{xz} = \max(S+1-N_x - Nz, 0)
\end{equation}

Each voxel in a subsweep can be computed in parallel. Mathematically, the $i^{th}$ subsweep from all directions can be computed in parallel. However, in practice, this results in a race condition on the GPU hardware. As such, parallelization in angular space is not implemented in the current version of DOCTORS.

\section{MCNP6 Generation}\label{sec:mcnpgen}
DOCTORS has the capability to generate MCNP6 input files from the CT mesh data and source specification provided by the user. The input file is procedurally generated using the information provided to DOCTORS. If insufficient information is provided, an error message is generated and no output file is made. Note that the output is always named "mcnp\_out.inp" so it will \textit{override} existing files. To save a file, it must be manually renamed to avoid accidental deletion!

The geometry is the largest section of the MCNP6 input file. The dimensions of the phantom mesh are used to generate planar surfaces at the appropriate locations. Those surfaces are then arranged into cells defining each voxel. The density information is pulled directly from the internally generated data built during the CT-number to material conversion.

The source is the most difficult component to generate. Currently, all source types except fan beams can be generated automatically. The difficulty in generating fan beams comes from an intrinsic limitation of MCNP6. MCNP6 can only model cone beams through a biasing mechanism; it cannot model other kinds of beams. In practice, this is rarely a major limitation since physical collimators can be added to a input model to generate the desired beam shape. However, since DOCTORS already has a geometry mesh, addition of collimators is very difficult since the collimator becomes "smeared" across voxels.

\section{Hardware}\label{sec:hardware}
For this work, a computer with an Intel i7-5960X 8 core (16 hyperthreads) processor with a base clock speed of 3.5 GHz and an Nvidia Titan Z graphics card was used. Currently, if the problem requires more memory than is available on the GPU, the problem can still be solved, but much more time will be required to to copy overhead between the CPU and GPU. If the problem requires more memory than either the GPU or CPU can provide, an error is thrown and the simulation is not run.

Only Nvidia GPUs are recognized by DOCTORS since it relies on the CUDA interface which is a Nvidia proprietary product. There are no limitations on the CPU except that it be Little-Endian based (nearly all major CPUs meet this requirment). Big-Endian hardware would fail to parse the data files correctly.

### 2017-07-05 docprep/chapintro.tex
Fixed typos and updated LaTeX template

Since its introduction in 1973 by \citet{ref:hounsfieldg} computed tomography (CT) has become pervasive in medical diagnostics as improved algorithms and techniques give doctors access to higher quality information. Increased information results in faster and more accurate diagnosis. However, the exponentially increasing usage of CT and dose patients receive from it has raised concerns of potential long term risks~\citep{ref:brennerd,ref:einsteina1,ref:abramsh,ref:einsteina2,ref:mccolloughc,ref:yul}. These concerns are exemplified by the push for low dose CT.

@@

In addition to rapidly computing the patient dose, another goal of DOCTORS is to present the code in a user-friendly fashion. To this end, a graphical user interface (GUI) was developed. The GUI was built using the Qt5 graphics framework for the windows, buttons, and other necessary widgets. The GUI leads the user through the steps necessary to use the code in an intuitive way by using colors to indicated required and completed steps. The output is then plotted graphically as well as sent to an ASCII text file for more advanced postprocessing. Utilization of Qt5 also makes the code portable, it can be compiled on either Linux or Windows operating systems with no changes to the code. More details about the GUI are included in Section~\ref{sec:gui}.

INSERT BRIEF DISCUSSION OF RESULTS

Another, similar, application is dose estimation of patients receiving radiation therapy. Often, before the procedure, a time dependent, 4D CT scan of the patient is taken so that radiologists can account for breathing patterns during administration of the treatment~\citep{ref:pant}. With further development, this methodology may enable real time dose computation as the treatment is administered. This would be greatly beneficial to both patients and the doctors administering the procedure.

>>>>

Since its introduction in 1973 by \citet{ref:hounsfieldg}, computed tomography (CT) has become pervasive in medical diagnostics as improved algorithms and techniques give doctors access to higher quality information. Increased information results in faster and more accurate diagnosis. However, the exponentially increasing usage of CT and dose patients receive from it has raised concerns of potential long term risks~\citep{ref:brennerd,ref:einsteina1,ref:abramsh,ref:einsteina2,ref:mccolloughc,ref:yul}. These concerns are exemplified by the push for low dose CT.

@@

In addition to rapidly computing the patient dose, another goal of DOCTORS is to present the code in a user-friendly fashion. To this end, a graphical user interface (GUI) was developed. The GUI was built using the Qt5 graphics framework for the windows, buttons, and other necessary widgets. The GUI leads the user through the steps necessary to use the code in an intuitive way by using colors to indicate required and completed steps. The output is then plotted graphically as well as sent to an ASCII text file for more advanced postprocessing. Utilization of Qt5 also makes the code portable, it can be compiled on either Linux or Windows operating systems with no changes to the code. More details about the GUI are included in Section~\ref{sec:gui}.

Overall, good qualitative agreement was found between the reference code, MCNP6, and DOCTORS. The primary source of error is believed to be the angular treatment of the transport which is causing the flux inside the patient to be underestimated while the flux at the very periphery is overestimated.

In the future, DOCOTRS may be extended to other application domains such as dose estimation of patients receiving radiation therapy. Often, before the procedure, a time dependent, 4D CT scan of the patient is taken so that radiologists can account for breathing patterns during administration of the treatment~\citep{ref:pant}. With further development, this methodology may enable real time dose computation as the treatment is administered. This would be greatly beneficial to both patients and the doctors administering the procedure.
